Natural Stories GAM analysis

### 2020-07-20 Analysis/nat_stories_gam.Rmd
with new gam

---
title: "Natural Stories LM analysis"
output: github_document
---
```{r, include=F}
knitr::opts_chunk$set(echo = FALSE, warning=F)
options(knitr.table.format = "html")
library(tidyverse)
library(readr)
library(brms)
library(lme4)
library(rstan)
library(tidybayes)
library(knitr)
library(mgcv)
library(mgcViz)
library(tidymv)
theme_set(theme_bw())
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

@@

data_error_summ <- data_filt %>% 
  mutate(correct.num=ifelse(correct=="yes", 1,0)) %>% 
  group_by(subject, critical, practice) %>%
  filter(type!="practice") %>% 
  filter(rt<5000) %>% 
  summarize(pct_correct=mean(correct.num)) %>% 
  ungroup() %>% 
  mutate(is.attentive=ifelse(pct_correct>.8, T,F)) %>% 
  select(subject, is.attentive)

@@

We sometimes center the predictors, but don't rescale them. (We center them for the brm models, but don't center surprisal for the GAM model).

The model formula we use is rt ~ surp* length +  freq * length + past_surp * past_length + past_freq * past_length.

@@

```{r}

lab_summary <- select_labs %>% summarize_at(c("length","freq","txl_surp","grnn_surp","ngram_surp"), list(mean=mean, stdev=sd, min=min, max=max)) %>% 
  pivot_longer(everything()) %>% arrange(name)

ggplot(select_labs, aes(length))+geom_bar()

@@

## BRM models

We include a by-subject effect for everything, and a by_word random intercept (full mixed effects).

Priors:

- normal(1000,1000) for intercept -- we think RTs are about 1 second usually
- normal(0,500) for beta and sd -- we don't really know what effects are
- lkj(1) for correlations -- we don't have reason to think correlations might go any particular way

```{r}
  priors <- c(
      set_prior("normal(1000, 1000)", class="Intercept"),
      set_prior("normal(0, 500)", class="b"),
      set_prior("normal(0, 500)", class="sd"),
      set_prior("lkj(1)",       class="cor"))

@@

### On pre-error data only 
```{r}
brm_ngram_interact <- brm(rt ~ ngram_center* length_center + freq_center * length_center + 
                         past_c_ngram*past_c_length + past_c_freq*past_c_length + 
                         (ngram_center* length_center + freq_center * length_center + 
                            past_c_ngram *past_c_length+ past_c_freq*past_c_length|subject)+
                         (1|Word_ID), data=labelled_pre_error, 
                       file="brm_ngram_interact", prior=priors, control=list(adapt_delta=.95))

brm_grnn_interact <- brm(rt ~ grnn_center * length_center+ freq_center * length_center + 
                         past_c_grnn *past_c_length+ past_c_freq*past_c_length + 
                         (grnn_center* length_center + freq_center * length_center + 
                            past_c_grnn*past_c_length + past_c_freq*past_c_length|subject)+
                        (1|Word_ID), data=labelled_pre_error, 
                       file="brm_grnn_interact", prior=priors, control=list(adapt_delta=.95))

brm_txl_interact <- brm(rt ~ txl_center* length_center + freq_center * length_center + 
                         past_c_txl*past_c_length + past_c_freq*past_c_length + 
                         (txl_center* length_center + freq_center * length_center + 
                            past_c_txl *past_c_length+ past_c_freq*past_c_length|subject)+
                       (1|Word_ID), data=labelled_pre_error, 
                       file="brm_txl_interact", prior=priors, control=list(adapt_delta=.95))

@@

```{r}

show_summary <- function(model){
  intervals <- gather_draws(model, `b_.*`, regex=T) %>% mean_qi()

  stats <- gather_draws(model, `b_.*`, regex=T) %>% 
    mutate(above_0=ifelse(.value>0, 1,0)) %>% 
    group_by(.variable) %>% 
    summarize(pct_above_0=mean(above_0)) %>% 
    mutate(`P` = signif(2*pmin(pct_above_0,1-pct_above_0), digits=2)) %>% 
    left_join(intervals, by=".variable") %>% 
    mutate(lower=round(.lower, digits=1),
           upper=round(.upper, digits=1),
           E=round(.value, digits=1),
           `CI`=str_c("[",lower,", ", upper,"]"),
           Term=str_sub(.variable, 3, -1),
           ) %>% 
    select(Term, `E`, `CI`,`P`)

  stats
}
a <- show_summary(brm_txl_interact) %>% mutate(model="TXL") %>% mutate(Term=str_replace(Term,"txl","surp"))
b <- show_summary(brm_grnn_interact) %>% mutate(model="GRNN") %>% mutate(Term=str_replace(Term,"grnn","surp"))
c <- show_summary(brm_ngram_interact) %>% mutate(model="5-gram") %>% mutate(Term=str_replace(Term,"ngram","surp"))

summ <- a %>% union(b) %>% union(c) %>% pivot_wider(names_from="model", values_from=c(`E`,`CI`,`P`)) %>% 
  select(Term, `E_5-gram`, `CI_5-gram`, `P_5-gram`, `E_GRNN`, CI_GRNN, P_GRNN, E_TXL, CI_TXL, P_TXL)

library(xtable)
print(xtable(summ,digits=c(0,0,1,0,2,1,0,2,1,0,2)), include.rownames=FALSE)
```

### On post-error as well

```{r}
brm_ngram_interact_post <- brm(rt ~ ngram_center* length_center + freq_center * length_center + 
                         past_c_ngram*past_c_length + past_c_freq*past_c_length + 
                         (ngram_center* length_center + freq_center * length_center + 
                            past_c_ngram *past_c_length+ past_c_freq*past_c_length|subject)+
                         (1|Word_ID), data=labelled_post_error, 
                       file="brm_ngram_interact_post", prior=priors, control=list(adapt_delta=.95))

brm_grnn_interact_post <- brm(rt ~ grnn_center * length_center+ freq_center * length_center + 
                         past_c_grnn *past_c_length+ past_c_freq*past_c_length + 
                         (grnn_center* length_center + freq_center * length_center + 
                            past_c_grnn*past_c_length + past_c_freq*past_c_length|subject)+
                        (1|Word_ID), data=labelled_post_error, 
                       file="brm_grnn_interact_post", prior=priors, control=list(adapt_delta=.95))

brm_txl_interact_post <- brm(rt ~ txl_center* length_center + freq_center * length_center + 
                         past_c_txl*past_c_length + past_c_freq*past_c_length + 
                         (txl_center* length_center + freq_center * length_center + 
                            past_c_txl *past_c_length+ past_c_freq*past_c_length|subject)+
                       (1|Word_ID), data=labelled_post_error, 
                       file="brm_txl_interact_post", prior=priors, control=list(adapt_delta=.95))

```
```{r}
show_summary_2 <- function(model){
  intervals <- gather_draws(model, `b_.*`, regex=T) %>% mean_qi()

  stats <- gather_draws(model, `b_.*`, regex=T) %>% 
    mutate(above_0=ifelse(.value>0, 1,0)) %>% 
    group_by(.variable) %>% 
    summarize(pct_above_0=mean(above_0)) %>% 
    mutate(`P-value equivalent` = signif(2*pmin(pct_above_0,1-pct_above_0), digits=2)) %>% 
    left_join(intervals, by=".variable") %>% 
    mutate(lower=round(.lower, digits=2),
           upper=round(.upper, digits=2),
           `Credible Interval`=str_c("[",lower,", ", upper,"]"),
           Term=str_sub(.variable, 3, -1),
           Estimate=round(.value, digits=2)) %>% 
    select(Term, Estimate, `Credible Interval`, `P-value equivalent`)

  stats
}
```

### Summaries

```{r}
summary(brm_ngram_interact_post)
summary(brm_ngram_interact)

summary(brm_grnn_interact_post)
summary(brm_grnn_interact)

summary(brm_txl_interact_post)
summary(brm_txl_interact)
```

I think there might be ways to have mixed effects for this, but for now we're not. Also, couldn't figure out good ways of getting graphs from the full models, so just fitting the pieces within ggplot for graphs.

```{r gams-2}
gam_ngram_1 <- gam(rt ~ s(ngram_surp, bs="cr", k=20)+ te(freq, length, bs="cr")+s(past_ngram, bs="cr", k=20)+te(past_freq, past_length, bs="cr"), data=labelled_pre_error)

gam_txl_1 <- gam(rt ~ s(txl_surp, bs="cr", k=20)+ te(freq, length, bs="cr")+s(past_txl, bs="cr", k=20)+te(past_freq, past_length, bs="cr"), data=labelled_pre_error)

gam_grnn_1 <- gam(rt ~ s(grnn_surp, bs="cr", k=20)+ te(freq, length, bs="cr")+s(past_grnn, bs="cr", k=20)+te(past_freq, past_length, bs="cr"), data=labelled_pre_error)

```{r}
for_graph <- labelled_pre_error %>% 
  select(rt, txl_surp, ngram_surp, grnn_surp, past_txl, past_grnn, past_ngram) %>%
  pivot_longer(cols=`txl_surp`:`past_ngram`,names_to="label", values_to="data") %>% 
  mutate(type=case_when(
         label %in% c("past_ngram", "past_txl", "past_grnn")~"Previous",
         T~"Current"),
         model=case_when(
           label %in% c("txl_surp", "past_txl")~"TXL",
           label %in% c("grnn_surp", "past_grnn")~"GRNN",
           T ~ "5-gram"
         ))

ggplot(for_graph, aes(x=data, y=rt))+
         geom_smooth(method="gam",formula= y~s(x, bs="cr", k=20),color="black")+
         facet_grid(type~model)+
        labs(x="Surprisal (bits)", y="Reaction Time (ms)")

Including post-error data.

```{r gams-3}
gam_ngram_2 <- gam(rt ~ s(ngram_surp, bs="cr", k=20)+ te(freq, length, bs="cr")+s(past_ngram, bs="cr", k=20)+te(past_freq, past_length, bs="cr"), data=labelled_post_error)

gam_txl_2 <- gam(rt ~ s(txl_surp, bs="cr", k=20)+ te(freq, length, bs="cr")+s(past_txl, bs="cr", k=20)+te(past_freq, past_length, bs="cr"), data=labelled_post_error)

gam_grnn_2 <- gam(rt ~ s(grnn_surp, bs="cr", k=20)+ te(freq, length, bs="cr")+s(past_grnn, bs="cr", k=20)+te(past_freq, past_length, bs="cr"), data=labelled_post_error)

```{r plot_gam-3}
for_graph_2 <- labelled_post_error %>% 
  select(rt, txl_surp, ngram_surp, grnn_surp, past_txl, past_grnn, past_ngram) %>%
  pivot_longer(cols=`txl_surp`:`past_ngram`,names_to="label", values_to="data") %>% 
  mutate(type=case_when(
         label %in% c("past_ngram", "past_txl", "past_grnn")~"Previous",
         T~"Current"),
         model=case_when(
           label %in% c("txl_surp", "past_txl")~"TXL",
           label %in% c("grnn_surp", "past_grnn")~"GRNN",
           T ~ "5-gram"
         ))

ggplot(for_graph_2, aes(x=data, y=rt))+
         geom_smooth(method="gam",formula= y~s(x, bs="cr", k=20),color="black")+
         facet_grid(type~model)+
        labs(x="Surprisal (bits)", y="Reaction Time (ms)")

ggsave("../Papers/gam2.pdf", width=3.5, height=2, unit="in")

for_graph_3 <- labelled_post_only %>% 
  select(rt, txl_surp, ngram_surp, grnn_surp, past_txl, past_grnn, past_ngram) %>%
  pivot_longer(cols=`txl_surp`:`past_ngram`,names_to="label", values_to="data") %>% 
  mutate(type=case_when(
         label %in% c("past_ngram", "past_txl", "past_grnn")~"Previous",
         T~"Current"),
         model=case_when(
           label %in% c("txl_surp", "past_txl")~"TXL",
           label %in% c("grnn_surp", "past_grnn")~"GRNN",
           T ~ "5-gram"
         ))

ggplot(for_graph_3, aes(x=data, y=rt))+
         geom_smooth(method="gam",formula= y~s(x, bs="cr", k=20),color="black")+
         facet_grid(type~model)+
        labs(x="Surprisal (bits)", y="Reaction Time (ms)")

ggsave("../Papers/gam3.pdf", width=3.5, height=2, unit="in")

# TODOs for future

In terms of surprisal analyses, the possible next steps are doing GAM analysis more robustly (with mixed effects, presenting results from the models) and doing DTSR.

We could also try looking at the specific syntactic features that are in the dataset (the point of natural stories was to have a bunch of idioms and weird syntax); we could maybe analyse those?

>>>>

---
title: "Natural Stories GAM analysis"
output: github_document
---
```{r, include=F}
knitr::opts_chunk$set(echo = FALSE, warning=F)
options(knitr.table.format = "html")
library(tidyverse)
library(readr)
library(brms)
library(lme4)
library(rstan)
library(tidybayes)
library(knitr)
library(mgcv)
library(mgcViz)
library(tidymv)
library(rsample)
theme_set(theme_bw())
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

@@

data_error_summ <- data_filt %>% 
  mutate(correct.num=ifelse(correct=="yes", 1,0)) %>% 
  group_by(subject) %>%
  filter(type!="practice") %>% 
  filter(rt<5000) %>% 
  summarize(pct_correct=mean(correct.num)) %>% 
  ungroup() %>% 
  mutate(is.attentive=ifelse(pct_correct>.8, T,F)) %>% 
  select(subject, is.attentive)

@@

For GAM models, we center length and frequency but not surprisal. We want surprisal interpretable, but we also will be plotting it (at least for the bootstrapping) at length and frequencies set to 0, so they need to be centered. (Not sure this last piece is actually true/matters).

@@

Does not have hierarchical effects.

```{r for gam}

ngram_data <- labelled_pre_error %>% select(rt, surprisal=ngram_surp, prev_surp=past_ngram, 
                                            freq=freq_center, len=length_center,
                                            prev_freq=past_c_freq, prev_len=past_c_length
                                            )

grnn_data <- labelled_pre_error %>% select(rt, surprisal=grnn_surp, prev_surp=past_grnn, 
                                            freq=freq_center, len=length_center,
                                            prev_freq=past_c_freq, prev_len=past_c_length
                                            )

txl_data <- labelled_pre_error %>% select(rt, surprisal=txl_surp, prev_surp=past_txl, 
                                            freq=freq_center, len=length_center,
                                            prev_freq=past_c_freq, prev_len=past_c_length
                                            )

@@

```{r gams-2}
gam_ngram_1 <- gam(rt ~ s(surprisal, bs="cr", k=20)+ te(freq, len, bs="cr")+s(prev_surp, bs="cr", k=20)+te(prev_freq, prev_len, bs="cr"), data=ngram_data, method="REML")

gam_grnn_1 <- gam(rt ~ s(surprisal, bs="cr", k=20)+ te(freq, len, bs="cr")+s(prev_surp, bs="cr", k=20)+te(prev_freq, prev_len, bs="cr"), data=grnn_data, method="REML")

gam_txl_1 <- gam(rt ~ s(surprisal, bs="cr", k=20)+ te(freq, len, bs="cr")+s(prev_surp, bs="cr", k=20)+te(prev_freq, prev_len, bs="cr"), data=txl_data, method="REML")

@@

```{r plot-gam}

a <- get_gam_predictions(model=gam_ngram_1, series=surprisal, series_length=100) %>% select(surprisal, rt, CI_upper, CI_lower) %>% unique() %>% mutate(model="5-gram", s="Current")
b <- get_gam_predictions(model=gam_ngram_1, series=prev_surp, series_length=100) %>% select(surprisal=prev_surp, rt, CI_upper, CI_lower) %>% unique() %>% mutate(model="5-gram", s="Previous")
c <- get_gam_predictions(model=gam_grnn_1, series=surprisal, series_length=100) %>% select(surprisal, rt, CI_upper, CI_lower) %>% unique() %>% mutate(model="GRNN", s="Current")
d <- get_gam_predictions(model=gam_grnn_1, series=prev_surp, series_length=100) %>% select(surprisal=prev_surp, rt, CI_upper, CI_lower) %>% unique() %>% mutate(model="GRNN", s="Previous")
e <- get_gam_predictions(model=gam_txl_1, series=surprisal, series_length=100) %>% select(surprisal, rt, CI_upper, CI_lower) %>% unique() %>% mutate(model="TXL", s="Current")
f <- get_gam_predictions(model=gam_txl_1, series=prev_surp, series_length=100) %>% select(surprisal=prev_surp, rt, CI_upper, CI_lower) %>% unique() %>% mutate(model="TXL", s="Previous")

all <- a %>% union(b) %>% union(c) %>% union(d) %>% union(e) %>% union(f)

ggplot(all, aes(x=surprisal, y=rt, ymin=CI_lower, ymax=CI_upper))+
  geom_line()+
  geom_ribbon(alpha=.3)+
  facet_grid(s~model)+
  labs(x="Surprisal (bits)", y="Reaction Time (ms)")

#confirm that this really is just a prettied up version of what the usual output is
#plot(gam_txl_1, seWithMean = TRUE, shift = coef(gam_txl_1)[1], select=1)

@@
